Epilepsy example

### 2022-11-09 src/epil/epil.Rmd
Some improved writing, and get the results for the Laplace marginal into table for beta

---
title: "Epilepsy example"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
bibliography: citations.bib
abstract: |
  **Background** @rue2009approximate use the epilespy example from the OpenBUGs manual as demonstration.

  **Task** We implement the epilepsy example in Stan, `R-INLA`, `TMB`, `glmmTMB`, `tmbstan` and `aghq`. We then compare the results of the inference methods.
---

@@

Section 5.2. of @rue2009approximate demonstrates the application of `R-INLA` to a generalised linear mixed model.
The data, previously included as an example in the OpenBUGs manual, concern the results of a clinical trial for a new epilepsy drug.
Patients $i = 1, \ldots, 59$ are each either assigned the new drug $\texttt{Trt}_i = 1$ or placebo $\texttt{Trt}_i = 0$.
Each patient makes four visits the clinic $j = 1, \ldots, 4$.
The observations $y_{ij}$ give the number of seizures of the $i$th person in the two weeks preceeding their $j$th visit to the clinic.
The covariates are age $\texttt{Age}_i$, baseline seizure counts $\texttt{Base}_i$ and an indicator for the final clinic visit $\texttt{V}_4$, which are all centered.
The observations are modelling using a Poisson distribution $y_{ij} \sim \text{Poisson}(e^{\eta_{ij}})$ with linear predictor 
\begin{align*}
  \eta_{ij}       &= \beta_0 + \beta_\texttt{Base} \log(\texttt{Baseline}_j / 4) + \beta_\texttt{Trt} \texttt{Trt}_i +
                     \beta_{\texttt{Trt} \times \texttt{Base}} \texttt{Trt}_i \times \log(\texttt{Baseline}_j / 4) \\ 
                  &+ \beta_\texttt{Age} \log(\texttt{Age}_i) + \beta_{\texttt{V}_4} {\texttt{V}_4}_j +
                     \epsilon_i + \nu_{ij}, \quad i=1:59, \quad j=1:4,
\end{align*}
where the prior on each of the regression parameters is $\mathcal{N}(0, 100^2)$.
The random effects are IID $\epsilon_i \sim \mathcal{N}(0, 1/\tau_\epsilon)$ and $\nu_{ij} \sim \mathcal{N}(0, 1/\tau_\nu)$ with precision priors $\tau_\epsilon, \tau_\nu \sim \Gamma(0.001, 0.001)$.

@@

Implement the model based upon a response variable of length $N \times J$, where $N$ is the number of patients and $J$ is the number of visits to the clinic, and a model matrix with $K$ predictors (including the intercept term):

@@

@@

First, Hamiltonian Monte Carlo via the No-U-Turn algorithm, with four Markov chains of `iter = 1000` samples, and `warmup = 100` burn-in:

```{r}
dat <- list(N = N, J = J, K = K, X = X, y = y, E = make_epsilon_matrix(N, J))

fit1 <- stan(
  "epil.stan",
  data = dat,
  chains = 4,
  warmup = 100,
  iter = 1000,
  control = list(adapt_delta = 0.95),
  refresh = 0
)
```

@@

Both $\tau_\epsilon$ and $\tau_\nu$ have the same $\Gamma(0.001, 0.001)$ prior.
In `R-INLA`, the precision is internally represented as log-precision, therefore we must set a `loggamma` prior:

@@

We can use various settings for the strategy, which controls the approximation used for the latent field posterior marginals:

@@

The objective function `obj$fn` and its gradient `obj$gn` are a function of only the parameters, which in this instance are the six regression coefficients $\beta$ together with the logarithms of $\tau_\epsilon$ and $\tau_\nu$.
This can be checked with `names(obj$par)`.

@@

Optimise `obj` using 1000 iterations of the the `nlminb` optimiser, passing in the starting values `start`, objective function `objective` and its derivative `gradient`:

@@

To be more thorough, we can get all the parameters obtained using the MCMC:

@@

Using just a few of them, we can see that the difference between the `TMB` and Stan objectives is a constant:

@@

@@

@@

Create a function to get the Laplace marginal for the $i$th element of $\beta$:

@@

Now we can compute the Laplace marginal for all `r dat$K` indices of $\beta$:

```{r}
fine_grid <- seq(-5, 5, length.out = 1000)

@@

@@

```{r}
xi_laplace_marginal <- function(i, opt_theta, finegrid) {
  #' To fill
}
```

Start with `i = 1` for the first marginal of $\beta$:

```{r}
i <- 1

#' I don't know if this is the right grid for this application...
fine_grid <- seq(-5, 5, length.out = 1000)

dat$toggle <- index_controller[i, "toggle"]
dat$i <- index_controller[i, "i"]

@@

ggplot(pdf_and_cdf, aes(x = theta, y = pdf)) +
  geom_line() +
  theme_minimal() +
  labs(x = "beta[1]", y = "Posterior")
```

@@

df <- cbind(stan1, inla2, inla3, inla4, tmb5, glmmtmb6, tmbstan7, aghq8) %>%
  as.data.frame() %>%
  mutate(type = gl(2, 1, 12, labels = c("Mean", "SD")))

beta_i <- function(i) { c(paste0("beta_", i), paste0("sd(beta_", i, ")")) }
rownames(df) <- c(sapply(0:5, beta_i))
colnames(df) <- c("Stan", "R-INLA_G", "R-INLA_SL", "R-INLA_L", "TMB", "glmmTMB", "tmbstan", "aghq", "type")

>>>>

---
title: "Epilepsy example"
author:
- name: Adam Howes
output:
  bookdown::html_document2:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
    df_print: paged
    code_folding: show
    theme: lumen
bibliography: citations.bib
abstract: |
  **Background** @rue2009approximate use the epilespy example from the OpenBUGs manual as demonstration of the integrated nested Laplace approximation method.

  **Task** We implement the epilepsy example in Stan, `R-INLA`, `TMB`, `glmmTMB`, `tmbstan`, `aghq`, and a custom Laplace marginal version of `aghq`. We then compare the results of the inference methods.
---

@@

Section 5.2. of @rue2009approximate demonstrates the application of `R-INLA` to a generalised linear mixed model.
The data, previously included as an example in the OpenBUGs manual, concern the results of a clinical trial for a new epilepsy drug.
Patients $i = 1, \ldots, 59$ are each either assigned the new drug $\texttt{Trt}_i = 1$ or placebo $\texttt{Trt}_i = 0$.
Each patient makes four visits the clinic $j = 1, \ldots, 4$.
The observations $y_{ij}$ give the number of seizures of the $i$th person in the two weeks preceeding their $j$th visit to the clinic.
The covariates are age $\texttt{Age}_i$, baseline seizure counts $\texttt{Base}_i$ and an indicator for the final clinic visit $\texttt{V}_4$, which are all centered.

The observations are modeleed using a Poisson distribution $y_{ij} \sim \text{Poisson}(e^{\eta_{ij}})$ with linear predictor 
\begin{align*}
  \eta_{ij}       &= \beta_0 + \beta_\texttt{Base} \log(\texttt{Baseline}_j / 4) + \beta_\texttt{Trt} \texttt{Trt}_i +
                     \beta_{\texttt{Trt} \times \texttt{Base}} \texttt{Trt}_i \times \log(\texttt{Baseline}_j / 4) \\ 
                  &+ \beta_\texttt{Age} \log(\texttt{Age}_i) + \beta_{\texttt{V}_4} {\texttt{V}_4}_j +
                     \epsilon_i + \nu_{ij}, \quad i=1:59, \quad j=1:4,
\end{align*}
where the prior on each of the regression parameters, including the intercept, is $\mathcal{N}(0, 100^2)$.
The random effects are IID $\epsilon_i \sim \mathcal{N}(0, 1/\tau_\epsilon)$ and $\nu_{ij} \sim \mathcal{N}(0, 1/\tau_\nu)$ with precision priors $\tau_\epsilon, \tau_\nu \sim \Gamma(0.001, 0.001)$.

@@

We will implement the model based upon a response variable of length $N \times J$, where $N$ is the number of patients and $J$ is the number of visits to the clinic, and a model matrix with $K$ predictors (including the intercept term):

@@

```{r}
dat <- list(N = N, J = J, K = K, X = X, y = y, E = make_epsilon_matrix(N, J))
```

@@

First, Hamiltonian Monte Carlo (HMC) via the No-U-Turn Sampler (NUTS) algorithm, with four Markov chains of `iter = 1000` samples, and `warmup = 100` burn-in.
Markov chain Monte Carlo (MCMC) like NUTS are exact as the number of samples tends to infinity.
We provide the `stan` function with `file = "epil.stan"` which refers to a C++ file implementing the log-posterior in the Stan probabilistic programming language:

```{r}
fit1 <- stan(
  file = "epil.stan",
  data = dat,
  chains = 4,
  warmup = 100,
  iter = 1000,
  control = list(adapt_delta = 0.95),
  refresh = 0
)
```

@@

`R-INLA` is an R package which implements integrated nested Laplace approximations.
Rather than specifying the log-posterior via C++, the package uses a formula interface.

We specify the $\Gamma(0.001, 0.001)$ prior on the random effects $\tau_\epsilon$ and $\tau_\nu$ below.
In `R-INLA`, the precision is internally represented as log-precision, therefore we must set a `loggamma` prior.

@@

We can use various settings (`"gaussian"`, `"simplified.laplace"` and `"laplace"`) for the strategy passed to `control.inla`, which controls the approximation used for the latent field posterior marginals:

@@

@@

The objective function `obj$fn` and its gradient `obj$gn` are a function of only the parameters, which in this instance are the logarithms of $\tau_\epsilon$ and $\tau_\nu$.
This can be checked with `names(obj$par)`.
Optimise `obj` using 1000 iterations of the the `nlminb` optimiser, passing in the starting values `start`, objective function `objective` and its derivative `gradient`:

@@

This look different!
Is this a problem?
We can evaluate the NLL at some of the other parameters explored during the MCMC:

@@

Using just a few of them, we can see that the difference between the `TMB` and Stan objectives is a constant, so for our purposes they are functionally the same:

@@

Ok, this next section is going to be a bit more involved.

@@

`epil_beta_index.cpp` is a `TMB` template where the `beta` parameter has been replaced by `beta_i` and `beta_minus_i`.
As well, there is an additional `DATA_INTEGER(i)` allowing us to pass in the index of `i` that we would like to obtain a full Laplace approximation to.

@@

We create a function to get the Laplace marginal for the $i$th element of $\beta$.
Some key things to note about this:
We use the `map` argument to fix the hyperparameters to their MAP estimates, which requires passing `factor(NA)` to `map`, as well as setting the initialisation parameters for the hyperparameters to be their MAP estimates.
Setting `random = c("beta_minus_i", "nu", "epsilon")` integrates out all of the random effects but `beta_i`.
Quadrature is performed using `aghq::aghq` to obtain a the normalised posterior marginal.

@@

Now we can compute the Laplace marginal for all `r dat$K` indices of $\beta$.

```{r}
fine_grid <- seq(-5, 5, length.out = 500)

@@

Ok, great, so we got the Laplace marginals for `beta`.
What about if we wanted to get them for all of the elements of the latent field?
One way we could do that is to create separate `TMB` templates for each named random effect (`beta`, `nu` and `epsilon`).
To see what this would look like, have a look at `epil_beta_index.cpp`, `epil_nu_index.cpp` and `epil_epsilon_index.cpp`.

What might be better is to do it in one `TMB` template.

@@

```{r}
xi_laplace_marginal <- function(i, opt_theta, finegrid) {
  dat$toggle <- index_controller[i, "toggle"]
  dat$i <- index_controller[i, "i"]

@@

  return(list("pdf_and_cdf" = pdf_and_cdf, "quad" = quad))
}
```

Now we can compute the Laplace marginal for all `r nrow(index_controller)` random effects of $(\beta, \nu, \epsilon)$.

```{r}
fine_grid <- seq(-5, 5, length.out = 500)

#' It's possible to get errors with the initial value not being finite
safe_xi_laplace_marginal <- function(...) {
  return(tryCatch(xi_laplace_marginal(...), error = function(e) NULL))
}

laplace_x <- lapply(1:nrow(index_controller), safe_xi_laplace_marginal, opt_theta = opt_theta, finegrid = fine_grid)

laplace_beta_summary <- lapply(head(laplace_x, n = dat$K), function(result) summary(result$quad)$summarytable) %>%
  bind_rows()

rownames(laplace_beta_summary) <- paste0("beta[", 1:dat$K, "]")
```

@@

adam9 <- as.vector(t(laplace_beta_summary[1:6, 1:2]))

df <- cbind(stan1, inla2, inla3, inla4, tmb5, glmmtmb6, tmbstan7, aghq8, adam9) %>%
  as.data.frame() %>%
  mutate(type = gl(2, 1, 12, labels = c("Mean", "SD")))

beta_i <- function(i) { c(paste0("beta_", i), paste0("sd(beta_", i, ")")) }
rownames(df) <- c(sapply(0:5, beta_i))
colnames(df) <- c("Stan", "R-INLA_G", "R-INLA_SL", "R-INLA_L", "TMB", "glmmTMB", "tmbstan", "aghq", "adam", "type")
